---
title: "Data Memo"
author: "Dylan Ma"
output:
    html_document:
      toc: true
      toc_float: true
      code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

## 1. Dataset Overview

**Overview**

The dataset was published on kaggle by Jesse Mostipak: [Hotel Booking Demand](https://www.kaggle.com/datasets/jessemostipak/hotel-booking-demand), The original author Antonio et al collected booking records from a Lisbon city hotel and a Algarve resort hotel in Portugal and published at (https://pmc.ncbi.nlm.nih.gov/articles/PMC6297060/.) The Kaggle version combines the two original datasets into a single CSV file including data from 2015 to 2017

The dataset includes 119,390 observations and 32 variables, including 1 response variable (is_canceled) and 31 raw candidate predictors.

```{r load-data}
hotels <- read_csv("data.csv",
                   na = c("", "NA", "NULL"))

dim(hotels)
glimpse(hotels)
```

**Variable types**:

The data contains the following types of continuous variables

- Numeric: `lead_time`, `adr`, `days_in_waiting_list`, `adults`, `children`, `babies`, `stays_in_weekend_nights`, `stays_in_week_nights`, `booking_changes`, `previous_cancellations`
- Binary: `is_repeated_guest`, `is_canceled` 
- Date-related: `arrival_date_year/month/week_number/day_of_month`; `reservation_status_date`

And 

- Categorical: `hotel`, `meal`, `country`, `market_segment`, `distribution_channel`, `reserved_room_type`, `assigned_room_type`, `deposit_type`, `customer_type`, `agent`, `company`


**Missing data**:

The CSV contains NA values in four variables.

- Since `company` is "NA" for 94.31% of the observations and contains many sparse ID levels, I plan to replace it with a binary indicator representing whether a company was involved in the booking.
- `children` has only four missing values, I will probably either delete the rows or input with the median value.
- `agent` is recorded as NA for only 13.69% of the observations, therefore I plan to treat agent as a categorical variable, retain the most frequent 15 agent IDs, and combine all other IDs into an “Other” category, and NA as NA.
- `country` has 488 missing values which is about 0.41% of the observations. These values will be labeled as “NA,” Same as agent, we will collect the most 15-20 frequent countries and treat others as "other"



## 2. Research Question

The response variable for this project is `is_canceled`, equals to 1 if a reservation was canceled and 0 otherwise. Out of 119,390 bookings, 44,224 were canceled, resulting in an overall cancellation rate of 37.04%.

The primary research question is Whether reservation and customer characteristics can be used to predict whether a hotel booking will be canceled?. Because the response variable has two possible classes, this question will be addressed using a binary classification approach. I expect `lead_time`, `deposit_type`, `previous_cancellations`, `market_segment`, `distribution_channel`, `adr` and `total_of_special_requests` to be especially useful because they represent booking condition, source, rates, and cancellation history.

The primary goal of the project is predictive. I will compare several classification models and determine which model most accurately predicts cancellations for new observations frop the testing set. Exploratory data analysis and model interpretation will also provide a descriptive understanding of the characteristics associated with cancellation.



## 3. Timeline

| Week | Tasks |
|---|---|
| 3 | Data cleaning & EDA (NA handling, duplicate investigation, outlier checks, predictor selection) |
| 4 | Train/test split, model recipe, and feature engineering | 
| 5 | Fit/tune logistic regression, regularization, decision tree, QDA/LDA (if works); initial model comparison |
| 6 | Random forest, Gradient boosting, any other models I am interesrted with (XGBoost/ Elastic Net), final model comparison, interpretation, finishing |



## 4. Questions and Concerns

1. The dataset contains 31,994 duplicate rows. A random split could place identical observations in both the training and test sets and inflate test performance. I am considering to remove the duplicate copied rows and retain one occurrence of each unique row, given that arrival date/cancelation status/any other variables are identical.